---
title: "EPPS 6302: Assignment 3"
subtitle: "Analysis of Twitter Data: Biden-Xi Summit"
format: 
  html:
    code-fold: true
    toc-depth: 2
---

## Introduction

This assignment aims to analyze Twitter data related to the President Biden and Xi summit in November 2021. We will explore public sentiment, popular hashtags, and user engagement through various text analysis techniques.

## Data Loading and Preparation

```{r}
library(quanteda)
library(quanteda.textmodels)
library(quanteda.textplots)
library(readr)
library(ggplot2)
library(knitr)
library(tidyverse)

# Load the data
summit <- read_csv("https://raw.githubusercontent.com/datageneration/datamethods/master/textanalytics/summit_11162021.csv")

# Display the data
kable(head(summit))
```

## Latent Semantic Analysis (LSA)

Latent Semantic Analysis helps us understand underlying themes.

```{r}
sum_twt = summit$text
toks = tokens(sum_twt)
sumtwtdfm <- dfm(toks)

# Perform LSA
sum_lsa <- textmodel_lsa(sumtwtdfm)

# Plot LSA results
lsa_dim1 <- sum_lsa$docs[,1]
lsa_dim2 <- sum_lsa$docs[,2]

plot(lsa_dim1, lsa_dim2, xlab = "LSA Dimension 1", ylab = "LSA Dimension 2", main = "LSA of Summit Tweets", pch = 20)

```

## Hashtag Analysis

Analyzing the most popular hashtags.

```{r}
tweet_dfm <- tokens(sum_twt, remove_punct = TRUE) %>%
  dfm()
tag_dfm <- dfm_select(tweet_dfm, pattern = "#*")
# Extract top 50 hashtags
top_hashtags <- topfeatures(tag_dfm, 25)

# Convert to data frame for plotting
df_hashtags <- data.frame(
  hashtag = names(top_hashtags),
  frequency = top_hashtags
)

# Plot the bar graph
ggplot(df_hashtags, aes(x = reorder(hashtag, frequency), y = frequency)) +
  geom_bar(stat = "identity") +
  coord_flip() +  # Flips the axes for easier reading
  labs(title = "Top 25 Hashtags", x = "Hashtag", y = "Frequency") +
  theme_minimal()

```

## Hashtag Network Analysis

Visualizing how different hashtags are related.

```{r}
tweet_dfm <- tokens(sum_twt, remove_punct = TRUE) %>%
  dfm()

# Create a network plot of hashtags
tag_fcm <- fcm(dfm_select(tweet_dfm, pattern = "#*"))
toptag <- names(topfeatures(dfm_select(tweet_dfm, pattern = "#*"), 50))

topgat_fcm <- fcm_select(tag_fcm, pattern = toptag)
textplot_network(topgat_fcm, min_freq = 50, edge_alpha = 0.8, edge_size = 5)

```

## User Mention Analysis

Analyzing the most frequently mentioned users.

```{r}
# Extract user mentions
user_dfm <- dfm_select(tweet_dfm, pattern = "@*")
topuser <- names(topfeatures(user_dfm, 50))
user_fcm <- fcm(user_dfm)
user_fcm <- fcm_select(user_fcm, pattern = topuser)
textplot_network(user_fcm, min_freq = 10, edge_color = "#e05248", edge_alpha = 0.8, edge_size = 5)

```
